---
title: "Human and Environmental Impact of Extreme Weather Events"
output: html_document
---

# Synopsis

(To follow completion of results)

# Data Processing

## Data loading 

Two packages are required. The dlpyr package is used to process the data, the 
ggplot2 package to display results, xtable to print tables.
```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(xtable)
```


The raw data is provided as a csv file, archived using bzip. An initial load of 
ten rows is made and the data investigated

```{r testLoad}
storm_test <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE, nrows = 10)
str(storm_test)
```

The headers provide suitable column names, but most variables will not be required. 
The full load omits those variables not required for the rest of the analysis.

```{r loadData, cache=TRUE}
cols <- c('NULL', NA, rep('NULL', 4), NA, NA, rep('NULL', 14), rep(NA, 6), rep('NULL', 9))

storm <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE,
                  colClasses = cols)
```

## Processing

### Data over time

This data covers a period of over 60 years. It is worth investigating how data 
gathering around weather events have changed over time. The BGN_DATE variable 
is used to identify the year in which the event occurred. 

```{r dataOverTime}
years <- as.Date(gsub("(\\d+\\/\\d+\\/\\d+)\\s0:00:00", "\\1", storm$BGN_DATE), format = "%m/%d/%Y")

storm$BGN_DATE <- years 
```

### Human and economic impact

Populate a data frame with the human impact for the whole of the United States, 
ordered by number of deaths and injuries per type of event, initially removing 
those events that did not result in fatalities
```{r humanImpact, cache=TRUE}
storm_hi_agg <- storm %>%
    filter(FATALITIES != 0) %>%
    group_by(EVTYPE) %>%
    summarise(FATALITIES = sum(FATALITIES,na.rm = T), 
              INJURIES = sum(INJURIES, na.rm = T)) %>%
    arrange(desc(FATALITIES), desc(INJURIES))
```

Similarly, produce a summary of the economic impact in terms of damage to 
property and to crops. This data is stored as pairs of {numerator, exponent} 
variables, with the exponents represented as numbers or single character 
suffixes of mixed case. These are converted to integer multipliers prior to 
summation. Once again, non-contributory events are discarded prior to further 
processing.
```{r economicImpact, cache=TRUE, warning=FALSE}

storm_cost_significant <- filter(storm, (PROPDMG != 0 & CROPDMG != 0))

# Replace duplicates caused by case variance
storm_cost_significant$PROPDMGEXP <- sapply(storm_cost_significant$PROPDMGEXP,toupper)
storm_cost_significant$CROPDMGEXP <- sapply(storm_cost_significant$CROPDMGEXP,toupper)

# Function to translate exponents to integers
expand <- function(x) {
        if (is.na(as.integer(x))) {
                return (1)
        } else {
                if (x == "K") {
                        return (1000)
                } else {
                        if (x == "M") {
                                return (10^6)
                        } else {
                                if (x == "B") {
                                        return (10^9)
                                } else {
                                        return(10^(as.integer(x)))
                                }
                        }
                }
        }
}

# Vectors are created containing the re-constitued costs
tpd <- with(storm_cost_significant, PROPDMG * mapply(expand, PROPDMGEXP))
tcd <- with(storm_cost_significant, CROPDMG * mapply(expand, CROPDMGEXP))

# These are then bound to a reduced set of variables
storm_ei <- cbind(storm_cost_significant, PropertyCost = tpd, CropCost = tcd)

# From this, aggregate costs are calulated
property_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Prop_Cost = sum(PropertyCost)) %>%
        arrange(desc(Prop_Cost))

crop_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Crop_Cost = sum(CropCost)) %>%
        arrange(desc(Crop_Cost))

# Finally the full cost of damage to property and crops is calculated
storm_ei_agg <- merge(property_cost, crop_cost)

storm_ei_agg <- storm_ei_agg %>%
        mutate(Total_Cost = Prop_Cost + Crop_Cost) %>%
        arrange(desc(Total_Cost))
```

### Impact by geographic location

The gross statistic for the whole country is broken down by state by looking at 
the most impactful individual events recorded in the period under study.

```{r humanImpactState, cache=TRUE}
storm_hi_state <- filter(storm, FATALITIES != 0)

storm_hi_state <- storm_hi_state %>%
        select(STATE, EVTYPE, FATALITIES, INJURIES) %>%
        group_by(STATE) %>%
        filter(FATALITIES == max(FATALITIES)) %>%
        arrange(STATE)
```


```{r economicImpactState, cache=TRUE}
storm_ei_state <- storm_ei %>%
        mutate(Total_Cost = PropertyCost + CropCost) %>%
        select(STATE, EVTYPE, Total_Cost) %>%
        group_by(STATE) %>%
        filter(Total_Cost == max(Total_Cost)) %>%
        arrange(STATE)
```

# Results

## Recorded events over time
This histogram shows how the number of recorded events increased over time, with 
a sharp jump in the mid 1990s. As later events predominate, it has been decided 
that the effects of inflation on the value of damage sustained will not be 
included in this study.

```{r histogramDates}
ggplot(storm, aes(BGN_DATE)) + geom_histogram(fill="white", colour="darkgreen") +
    labs(title = "Figure 1. Distribution of recorded events over time", 
         x = "Years", y = "Count")
```

## Gross Impact

```{r humanImpactResults, results='asis'}
hi_tbl <- xtable(as.data.frame(storm_hi_agg[1:10,]), 
                 caption = "Table 1. Human impact of weather events")
print(hi_tbl, type ="HTML", html.table.attributes = "border = '0'")
```

```{r economicImpactResults, results='asis'}
ei_tbl <- xtable(as.data.frame(storm_ei_agg[1:10,]), 
                 caption = "Table 2. Economic impact of weather events")
print(ei_tbl, type ="HTML", html.table.attributes = "border = '0'")
```

## Local impact

The human impact per state of individual events can be seen in figure 1.

```{r humanImpactStateFig}
hi_states_gg <- ggplot(storm_hi_state, aes(STATE, FATALITIES, fill = EVTYPE)) +
    geom_bar(position="dodge", stat = "identity") +
    theme(axis.text.x = element_text(angle = 270)) +
    guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.5)) +
    labs(title = "Figure 2. Highest fatality event per state")
hi_states_gg
```

Tornados are predominant across a number of states, though the death toll from a 
single heat wave in Illinois stands out.

```{r economicImpactStatesFig}
ei_states_gg <- ggplot(storm_ei_state, aes(STATE, Total_Cost, fill = EVTYPE)) +
    geom_bar(position="dodge", stat = "identity") +
    theme(axis.text.x = element_text(angle = 270)) + scale_y_sqrt() + 
    guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.5)) +
    labs(title = "Figure 3. Most costly event per state")
# Needs a log scale
ei_states_gg
```


It should be noted that the second most devastating event across the country is 
spread across most states. There is no obvious peak state for 'EXCESSIVE HEAT', 
though it has killed more than 1900 people in the period 1950 - 2011.

The high numbers of people killed by inidivdual heat-related events in Illinois 
warrants further investigation.

It could be that deaths are more likely to be 
associated with a heat wave when one occurs, give
suggests that 
deaths that happen during a heat wave are automatically associ
```{r excessiveHeatState}
exc_heat <- storm[storm$EVTYPE == "EXCESSIVE HEAT" & storm$FATALITIES != 0, 
                  c("STATE", "FATALITIES")]

exc_heat_state <- exc_heat %>%
group_by(STATE) %>%
summarise(sum(FATALITIES))

```

