---
title: "Human and Environmental Impact of Extreme Weather Events"
output: html_document
---

Synopsis
========

(To follow completion of results)

Data Processing
===============
Load libraries required.
```{r libraries, message=FALSE}
library(dplyr)
```


Load the raw data.
```{r loadData, cache=TRUE}
storm <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE)
```

Populate a data frame with the human impact for the whole of the United States, 
ordered by number of deaths and injuries per type of event, initially removing 
those events that do not contribute to totals
```{r humanImpact, cache=TRUE}
storm_hi_agg <- storm %>%
    filter(FATALITIES != 0 & INJURIES != 0) %>%
    group_by(EVTYPE) %>%
    summarise(FATALITIES = sum(FATALITIES,na.rm = T), 
              INJURIES = sum(INJURIES, na.rm = T)) %>%
    arrange(desc(FATALITIES), desc(INJURIES))
```

Similarly, produce a summary of the economic impact in terms of damage to 
property and to crops. This data is stored as pairs of {numerator, exponent} 
variables, with the exponents represented as numbers or single character 
suffixes of mixed case. These are converted to integer multipliers prior to 
summation. Once again, non-contributory events are discarded prior to further 
processing.
```{r economicImpact, cache=TRUE, warning=FALSE}

storm_cost_significant <- filter(storm, (PROPDMG != 0 & CROPDMG != 0))

# Replace duplicates caused by case variance
storm_cost_significant$PROPDMGEXP <- sapply(storm_cost_significant$PROPDMGEXP,toupper)
storm_cost_significant$CROPDMGEXP <- sapply(storm_cost_significant$CROPDMGEXP,toupper)

# Function to translate exponents to integers
expand <- function(x) {
        if (is.na(as.integer(x))) {
                return (1)
        } else {
                if (x == "K") {
                        return (1000)
                } else {
                        if (x == "M") {
                                return (10^6)
                        } else {
                                if (x == "B") {
                                        return (10^9)
                                } else {
                                        return(10^(as.integer(x)))
                                }
                        }
                }
        }
}

# Vectors are created containing the re-constitued costs
tpd <- with(storm_cost_significant, PROPDMG * mapply(expand, PROPDMGEXP))
tcd <- with(storm_cost_significant, CROPDMG * mapply(expand, CROPDMGEXP))

# These are then bound to a reduced set of variables
storm_ei <- cbind(storm_cost_significant[c(1,2,7,8,25:28)], 
                        PropertyCost = tpd, CropCost = tcd)

# From this, aggregate costs are calulated
property_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Prop_Cost = sum(PropertyCost)) %>%
        arrange(desc(Prop_Cost))

crop_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Crop_Cost = sum(CropCost)) %>%
        arrange(desc(Crop_Cost))

# Finally the full cost of damage to property and crops is calculated
storm_ei_agg <- merge(property_cost, crop_cost)

storm_ei_agg <- storm_ei_agg %>%
        mutate(Total_Cost = Prop_Cost + Crop_Cost) %>%
        arrange(desc(Total_Cost))
```

To do:

 - Aggregate similar and misnamed events
 - Breakdown top impactful events by geographic location, probably state
 - Map impact per state
 
Results
=======