---
title: "Human and Environmental Impact of Extreme Weather Events"
output: html_document
---

Synopsis
========

(To follow completion of results)

Data Processing
===============
Load libraries required.
```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
```


Load the raw data.
```{r loadData, cache=TRUE}
storm <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE)
```

Populate a data frame with the human impact for the whole of the United States, 
ordered by number of deaths and injuries per type of event, initially removing 
those events that did not result in fatalities
```{r humanImpact, cache=TRUE}
storm_hi_agg <- storm %>%
    filter(FATALITIES != 0) %>%
    group_by(EVTYPE) %>%
    summarise(FATALITIES = sum(FATALITIES,na.rm = T), 
              INJURIES = sum(INJURIES, na.rm = T)) %>%
    arrange(desc(FATALITIES), desc(INJURIES))
```

Similarly, produce a summary of the economic impact in terms of damage to 
property and to crops. This data is stored as pairs of {numerator, exponent} 
variables, with the exponents represented as numbers or single character 
suffixes of mixed case. These are converted to integer multipliers prior to 
summation. Once again, non-contributory events are discarded prior to further 
processing.
```{r economicImpact, cache=TRUE, warning=FALSE}

storm_cost_significant <- filter(storm, (PROPDMG != 0 & CROPDMG != 0))

# Replace duplicates caused by case variance
storm_cost_significant$PROPDMGEXP <- sapply(storm_cost_significant$PROPDMGEXP,toupper)
storm_cost_significant$CROPDMGEXP <- sapply(storm_cost_significant$CROPDMGEXP,toupper)

# Function to translate exponents to integers
expand <- function(x) {
        if (is.na(as.integer(x))) {
                return (1)
        } else {
                if (x == "K") {
                        return (1000)
                } else {
                        if (x == "M") {
                                return (10^6)
                        } else {
                                if (x == "B") {
                                        return (10^9)
                                } else {
                                        return(10^(as.integer(x)))
                                }
                        }
                }
        }
}

# Vectors are created containing the re-constitued costs
tpd <- with(storm_cost_significant, PROPDMG * mapply(expand, PROPDMGEXP))
tcd <- with(storm_cost_significant, CROPDMG * mapply(expand, CROPDMGEXP))

# These are then bound to a reduced set of variables
storm_ei <- cbind(storm_cost_significant[c(1,2,7,8,25:28)], 
                        PropertyCost = tpd, CropCost = tcd)

# From this, aggregate costs are calulated
property_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Prop_Cost = sum(PropertyCost)) %>%
        arrange(desc(Prop_Cost))

crop_cost <- storm_ei %>%
        group_by(EVTYPE) %>%
        summarise(Crop_Cost = sum(CropCost)) %>%
        arrange(desc(Crop_Cost))

# Finally the full cost of damage to property and crops is calculated
storm_ei_agg <- merge(property_cost, crop_cost)

storm_ei_agg <- storm_ei_agg %>%
        mutate(Total_Cost = Prop_Cost + Crop_Cost) %>%
        arrange(desc(Total_Cost))
```

The gross statistic for the whole country is further broken down by state.

```{r humanImpactState}
storm_hi_state <- filter(storm, FATALITIES != 0)

storm_hi_state <- storm_hi_state %>%
        select(STATE, EVTYPE, FATALITIES, INJURIES) %>%
        group_by(STATE) %>%
        filter(FATALITIES == max(FATALITIES)) %>%
        arrange(STATE)
```


```{r economicImpactState}
storm_ei_state <- storm_ei %>%
        mutate(Total_Cost = PropertyCost + CropCost) %>%
        select(STATE, EVTYPE, Total_Cost) %>%
        group_by(STATE) %>%
        filter(Total_Cost == max(Total_Cost)) %>%
        arrange(STATE)
```

To do:

 - Aggregate similar and misnamed events
 - Map impact per state
 
Results
=======


The human impact per state can be seen in figure X (Caption needed)

```{r humanImpactStateFig}
hi_states_gg <- ggplot(storm_hi_state, aes(STATE, FATALITIES, fill = EVTYPE)) +
        geom_bar(position="dodge", stat = "identity") +
        theme(axis.text.x = element_text(angle = 270)) +
        guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.5))
hi_states_gg
```

```{r economicImpactStatesFig}
ei_states_gg <- ggplot(storm_ei_state, aes(STATE, Total_Cost, fill = EVTYPE)) +
        geom_bar(position="dodge", stat = "identity") +
        theme(axis.text.x = element_text(angle = 270)) +
        guides(fill = guide_legend(keyheight = 0.8, keywidth = 0.5))
ei_states_gg
```

It should be noted that the second most devastating event across the country is 
spread across most states. There is no obvious peak state for 'EXCESSIVE HEAT', 
though it has killed more than 1900 people in the period 1950 - 2011.

The high numbers of people killed by inidivdual heat-related events in Illinois 
warrants further investigation.

It could be that deaths are more likely to be 
associated with a heat wave when one occurs, give
suggests that 
deaths that happen during a heat wave are automatically associ
```{r excessiveHeatState}
exc_heat <- storm[storm$EVTYPE == "EXCESSIVE HEAT" & storm$FATALITIES != 0, 
                  c("STATE", "FATALITIES")]

exc_heat_state <- exc_heat %>%
group_by(STATE) %>%
summarise(sum(FATALITIES))

```

